---
title: "Movie Recommendation System: KNN vs Matrix Factorization"
author: "Javier Kangditya"
output: html_document
---

# Introduction

## Project Overview
The goal of this project is to build and compare two different recommendation systems for movies based on user rating behavior. The models used in this project are **K-Nearest Neighbors (KNN)** and **Matrix Factorization (SVD)**. The objective is to understand how each method performs in predicting user preferences and generating personalized movie recommendations.

## Dataset Description
The dataset is obtained from MovieLens and contains two main files used in this project:

- **rating.csv**  
  Contains user-movie interactions with the following features:
  - userId  
  - movieId  
  - rating  
  - timestamp  

- **movie.csv**  
  Contains movie metadata including:
  - movieId  
  - title  
  - genres  

The dataset represents a user-item rating matrix that will be used for collaborative filtering techniques.

## Project Goals
1. Load and preprocess movie rating data  
2. Build a user-item matrix for recommendation  
3. Implement K-Nearest Neighbors (KNN) based recommender system  
4. Implement Matrix Factorization (SVD) based recommender system  
5. Evaluate and compare both models  
6. Generate movie recommendations for users  
7. Provide insights on which method performs better for sparse rating data

---

# 1. Data Cleaning
```{r load_and_clean, message=FALSE, warning=FALSE}
# =========================================================
# 1. LIBRARIES
# =========================================================
library(tidyverse)
library(data.table)

# =========================================================
# 2. LOAD RAW DATA
# =========================================================
ratings_raw <- fread("Data/Raw/rating.csv")
movies_raw  <- fread("Data/Raw/movie.csv")

# RAW PREVIEW (LIGHTWEIGHT)
cat("===== RAW DATA OVERVIEW =====\n")
cat("Ratings dim:\n")
print(dim(ratings_raw))

cat("\nMovies dim:\n")
print(dim(movies_raw))

cat("\nSample Ratings:\n")
print(head(ratings_raw, 5))

cat("\nSample Movies:\n")
print(head(movies_raw, 5))

# =========================================================
# 3. CLEAN DATA
# =========================================================
ratings <- unique(na.omit(ratings_raw))
movies  <- unique(na.omit(movies_raw))

# Type conversion (FAST data.table style)
ratings[, userId := as.integer(userId)]
ratings[, movieId := as.integer(movieId)]
ratings[, rating  := as.numeric(rating)]

movies[, movieId := as.integer(movieId)]
movies[, title   := as.character(title)]
movies[, genres  := as.character(genres)]

# =========================================================
# 4. FILTER SPARSE DATA 
# =========================================================
user_counts  <- ratings[, .N, by = userId]
movie_counts <- ratings[, .N, by = movieId]

ratings <- ratings[
  userId %in% user_counts[N >= 10, userId] &
  movieId %in% movie_counts[N >= 5, movieId]
]

# =========================================================
# 5. FINAL CLEANED CHECK
# =========================================================
cat("\n===== CLEANED DATA OVERVIEW =====\n")
cat("Ratings dim:\n")
print(dim(ratings))

cat("\nMovies dim:\n")
print(dim(movies))

# =========================================================
# 6. SAVE PROCESSED DATA
# =========================================================
saveRDS(ratings, "Data/Processed/ratings_clean.rds")
saveRDS(movies, "Data/Processed/movies_clean.rds")

# =========================================================
# 7. LOAD & PREVIEW PROCESSED DATA
# =========================================================
ratings <- readRDS("Data/Processed/ratings_clean.rds")
movies  <- readRDS("Data/Processed/movies_clean.rds")

cat("\n===== PROCESSED DATA OVERVIEW =====\n")

cat("Processed Ratings dim:\n")
print(dim(ratings))

cat("\nProcessed Movies dim:\n")
print(dim(movies))

cat("\nProcessed Ratings Sample:\n")
print(head(ratings, 5))

cat("\nProcessed Movies Sample:\n")
print(head(movies, 5))
```

---

# 2. Exploratory Data Analysis 
```{r eda, message=FALSE, warning=FALSE}

library(tidyverse)

# =========================================================
# LOAD DATA
# =========================================================
ratings <- readRDS("Data/Processed/ratings_clean.rds")
movies  <- readRDS("Data/Processed/movies_clean.rds")

# =========================================================
# GLOBAL CLEAN (IMPORTANT FIX)
# =========================================================
ratings <- ratings %>%
  filter(!is.na(userId), !is.na(movieId), !is.na(rating))

# =========================================================
# 1. DATA SUMMARY
# =========================================================
cat("===== DATA SUMMARY =====\n")
cat("Total Users:", n_distinct(ratings$userId), "\n")
cat("Total Movies:", n_distinct(ratings$movieId), "\n")
cat("Total Ratings:", nrow(ratings), "\n")

# =========================================================
# 2. RATING DISTRIBUTION
# =========================================================
p1 <- ggplot(ratings, aes(x = rating)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Rating Distribution", x = "Rating", y = "Count")

ggsave("Outputs/Figures/01_rating_distribution.png", p1, width = 6, height = 4)
print(p1)

# =========================================================
# 3. USER ACTIVITY
# =========================================================
user_activity <- ratings %>% count(userId)

p2 <- ggplot(user_activity, aes(x = n)) +
  geom_histogram(binwidth = 10, fill = "darkgreen", color = "black") +
  theme_minimal() +
  labs(title = "User Activity Distribution", x = "Number of Ratings", y = "Users")

ggsave("Outputs/Figures/02_user_activity.png", p2, width = 6, height = 4)
print(p2)

# =========================================================
# 4. USER ACTIVITY VS AVG RATING
# =========================================================
user_stats <- ratings %>%
  group_by(userId) %>%
  summarise(
    n_ratings = n(),
    avg_rating = mean(rating),
    .groups = "drop"
  )

p3 <- ggplot(user_stats, aes(x = n_ratings, y = avg_rating)) +
  geom_point(alpha = 0.3) +
  theme_minimal() +
  labs(title = "User Activity vs Average Rating",
       x = "Number of Ratings",
       y = "Average Rating")

ggsave("Outputs/Figures/03_user_activity_vs_rating.png", p3, width = 7, height = 5)
print(p3)

# =========================================================
# 5. MOVIE POPULARITY
# =========================================================
movie_popularity <- ratings %>%
  count(movieId) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  left_join(movies, by = "movieId")

cat("\n===== TOP 10 MOST RATED MOVIES =====\n")
print(movie_popularity[, c("title", "n")])

write.csv(movie_popularity, "Outputs/Tables/movie_popularity.csv", row.names = FALSE)

# =========================================================
# 6. TOP RATED MOVIES
# =========================================================
top_rated <- ratings %>%
  group_by(movieId) %>%
  summarise(
    avg_rating = mean(rating),
    count = n(),
    .groups = "drop"
  ) %>%
  filter(count >= 20) %>%
  arrange(desc(avg_rating)) %>%
  head(10) %>%
  left_join(movies, by = "movieId")

p4 <- ggplot(top_rated, aes(x = reorder(title, avg_rating), y = avg_rating)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top Rated Movies", x = "Movie", y = "Average Rating")

ggsave("Outputs/Figures/04_top_rated_movies.png", p4, width = 7, height = 5)
print(p4)

# =========================================================
# 7. SPARSITY 
# =========================================================
n_users  <- as.numeric(n_distinct(ratings$userId))
n_movies <- as.numeric(n_distinct(ratings$movieId))
n_ratings <- as.numeric(nrow(ratings))

cat("DEBUG TYPES:\n")
cat("n_users:", n_users, typeof(n_users), "\n")
cat("n_movies:", n_movies, typeof(n_movies), "\n")
cat("n_ratings:", n_ratings, typeof(n_ratings), "\n")

# HARD CHECK
if (any(is.na(c(n_users, n_movies, n_ratings)))) {
  stop("NA detected in counts — data issue")
}

total_possible <- n_users * n_movies

cat("Total possible:", total_possible, "\n")

if (is.na(total_possible) || total_possible <= 0) {
  stop("Invalid total_possible")
}

sparsity <- 1 - (n_ratings / total_possible)

cat("\n===== SPARSITY =====\n")
cat("Sparsity:", round(sparsity, 8), "\n")

# =========================================================
# SAVE SUMMARY
# =========================================================
summary_table <- data.frame(
  Users = n_users,
  Movies = n_movies,
  Ratings = nrow(ratings),
  Sparsity = sparsity
)

write.csv(summary_table, "Outputs/Tables/eda_summary.csv", row.names = FALSE)
```

---

# 3. User-Item Matrix & Data Splitting
```{r matrix_split, message=FALSE, warning=FALSE}
# =========================================================
# LOAD LIBRARIES
# =========================================================
library(recommenderlab)
library(Matrix)
library(tidyverse)

# =========================================================
# LOAD CLEAN DATA
# =========================================================
ratings <- readRDS("Data/Processed/ratings_clean.rds")

# =========================================================
# GLOBAL CLEAN 
# =========================================================
ratings <- ratings %>%
  filter(!is.na(userId), !is.na(movieId), !is.na(rating))

# =========================================================
# CREATE USER–ITEM MATRIX (SPARSE)
# =========================================================

# Convert userId & movieId to index
user_index  <- as.numeric(factor(ratings$userId))
movie_index <- as.numeric(factor(ratings$movieId))

# Build sparse matrix
rating_matrix <- sparseMatrix(
  i = user_index,
  j = movie_index,
  x = ratings$rating
)

cat("Matrix Dimension:\n")
cat("Users:", nrow(rating_matrix), "\n")
cat("Movies:", ncol(rating_matrix), "\n")

# =========================================================
# CONVERT TO recommenderlab FORMAT
# =========================================================
rating_rrm <- as(rating_matrix, "realRatingMatrix")

cat("\nRealRatingMatrix Summary:\n")
rating_rrm

# =========================================================
# TRAIN–TEST SPLIT
# =========================================================
set.seed(123)

eval_scheme <- evaluationScheme(
  data = rating_rrm,
  method = "split",
  train = 0.8,
  given = 10,
  goodRating = 4
)

cat("\nEvaluation Scheme Created\n")

# =========================================================
# EXTRACT TRAIN & TEST DATA
# =========================================================
train_data <- getData(eval_scheme, "train")
known_data <- getData(eval_scheme, "known")
unknown_data <- getData(eval_scheme, "unknown")

cat("\nTrain/Test Split Summary:\n")
train_data
known_data
unknown_data

# =========================================================
# SAVE OBJECTS 
# =========================================================
saveRDS(rating_rrm, "Data/Processed/rating_matrix.rds")
saveRDS(eval_scheme, "Data/Processed/eval_scheme.rds")
```

---

# 4. Modeling and Recommendation System
```{r modeling, message=FALSE, warning=FALSE}
library(recommenderlab)

# =========================================================
# 1. LOAD DATA
# =========================================================
eval_scheme <- readRDS("Data/Processed/eval_scheme.rds")

train <- getData(eval_scheme, "train")
known   <- getData(eval_scheme, "known")
unknown <- getData(eval_scheme, "unknown")

n <- min(100, nrow(known))

known_small   <- known[1:n, ]
unknown_small <- unknown[1:n, ]

# =========================================================
# 2. MODELS
# =========================================================

# POPULAR
model_pop <- Recommender(train, method = "POPULAR")

# MATRIX FACTORIZATION (LIBMF / fallback SVD)
model_mf <- tryCatch({
  Recommender(train, method = "LIBMF",
              parameter = list(dim = 10, iter = 30, lambda = 0.01))
}, error = function(e) {
  Recommender(train, method = "SVD", parameter = list(k = 10))
})

# KNN (IBCF - reduced size for stability)
train_small <- train[1:min(500, nrow(train)), ]

model_knn <- Recommender(
  train_small,
  method = "IBCF",
  parameter = list(k = 30)
)

# =========================================================
# 3. PREDICTIONS
# =========================================================
pred_pop <- predict(model_pop, known_small, n = 10)
pred_mf  <- predict(model_mf,  known_small, n = 10)
pred_knn <- predict(model_knn, known_small, n = 10)

# =========================================================
# 4. GROUND TRUTH 
# =========================================================
get_actual <- function(test_data) {

  lapply(1:nrow(test_data), function(i) {

    user <- as(test_data[i, ], "list")[[1]]

    if (length(user) == 0) return(character(0))

    names(user[user > 0])
  })
}

actual <- lapply(1:nrow(unknown_small), function(i) {

  user <- as(unknown_small[i, ], "list")[[1]]

  if (length(user) == 0) return(character(0))

  names(user)
})

# =========================================================
# 5. METRICS
# =========================================================

recall_at_k <- function(pred, actual, k = 10) {

  pred_list <- as(pred, "list")
  scores <- c()

  for (i in seq_along(pred_list)) {

    if (i > length(actual)) break

    p <- pred_list[[i]][1:k]
    a <- actual[[i]]

    if (length(a) == 0) next

    hits <- length(intersect(p, a))
    scores <- c(scores, hits / length(a))
  }

  if (length(scores) == 0) return(0)
  mean(scores)
}

map_at_k <- function(pred, actual, k = 10) {

  pred_list <- as(pred, "list")
  ap <- c()

  for (i in seq_along(pred_list)) {

    if (i > length(actual)) break

    p <- pred_list[[i]][1:k]
    a <- actual[[i]]

    if (length(a) == 0) next

    hit <- 0
    sum_prec <- 0

    for (j in seq_along(p)) {
      if (p[j] %in% a) {
        hit <- hit + 1
        sum_prec <- sum_prec + hit / j
      }
    }

    if (hit > 0) {
      ap <- c(ap, sum_prec / min(length(a), k))
    }
  }

  if (length(ap) == 0) return(0)
  mean(ap)
}

dcg <- function(rel) {
  sum(rel / log2(seq_along(rel) + 1))
}

ndcg_at_k <- function(pred, actual, k = 10) {

  pred_list <- as(pred, "list")
  scores <- c()

  for (i in seq_along(pred_list)) {

    if (i > length(actual)) break

    p <- pred_list[[i]][1:k]
    a <- actual[[i]]

    if (length(a) == 0) next

    rel <- as.numeric(p %in% a)
    ideal <- sort(rel, decreasing = TRUE)

    dcg_val <- dcg(rel)
    idcg_val <- dcg(ideal)

    if (idcg_val == 0) next

    scores <- c(scores, dcg_val / idcg_val)
  }

  if (length(scores) == 0) return(0)
  mean(scores)
}

# =========================================================
# 6. EVALUATION RESULTS
# =========================================================
results <- data.frame(
  model = c("POPULAR", "MF", "KNN (IBCF)"),

  recall_at_10 = c(
    recall_at_k(pred_pop, actual),
    recall_at_k(pred_mf, actual),
    recall_at_k(pred_knn, actual)
  ),

  map_at_10 = c(
    map_at_k(pred_pop, actual),
    map_at_k(pred_mf, actual),
    map_at_k(pred_knn, actual)
  ),

  ndcg_at_10 = c(
    ndcg_at_k(pred_pop, actual),
    ndcg_at_k(pred_mf, actual),
    ndcg_at_k(pred_knn, actual)
  )
)

print(results)

# =========================================================
# 7. SAVE OUTPUT
# =========================================================
saveRDS(model_pop, "Outputs/Models/popular.rds")
saveRDS(model_mf,  "Outputs/Models/mf.rds")
saveRDS(model_knn, "Outputs/Models/knn_ibcf.rds")

write.csv(results,
          "Outputs/Tables/model_results.csv",
          row.names = FALSE)
```

---

# 5. MODEL COMPARISON VISUALIZATION
```{r visualization, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

# =========================================================
# 1. LOAD RESULTS 
# =========================================================
# results <- read.csv("Outputs/Tables/model_results.csv")

# =========================================================
# 2. RESHAPE DATA 
# =========================================================
results_long <- results %>%
  pivot_longer(
    cols = -model,
    names_to = "metric",
    values_to = "value"
  )

# =========================================================
# 3. VISUALIZATION
# =========================================================
plot_model <- ggplot(results_long,
                     aes(x = model, y = value, fill = model)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~metric, scales = "free") +
  theme_minimal() +
  labs(
    title = "Model Performance Comparison",
    x = "Model",
    y = "Score"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14, face = "bold")
  )

print(plot_model)

# =========================================================
# 4. SAVE PLOT
# =========================================================
ggsave(
  "Outputs/Figures/model_comparison.png",
  plot_model,
  width = 8,
  height = 5
)
```

---

# Conclusion & Insights

## 1. Summary of Findings
- **Recommendation System Evaluation:**  
  - Three models were evaluated: POPULAR, Matrix Factorization (MF), and K-Nearest Neighbors (KNN).  
  - Evaluation metrics used were Recall@10, MAP@10, and NDCG@10 to measure both retrieval quality and ranking performance.  
  - The dataset shows a strong popularity bias, where a small number of frequently rated items dominate user preferences.  

- **Model Performance:**  
  - The POPULAR model achieved the best overall performance across all metrics, especially in MAP@10 and NDCG@10.  
  - The MF model showed moderate performance but was limited by data sparsity and insufficient interaction density.  
  - The KNN (IBCF) model performed the weakest, indicating that user-item similarity is not stable in sparse conditions.  

- **Evaluation Insights:**  
  - Recall@10 values are relatively low, which is expected in recommender systems with a large item space.  
  - MAP@10 and NDCG@10 provide better insight into ranking quality, showing that POPULAR consistently ranks relevant items higher.  
  - Overall, simpler baseline models can outperform more complex models depending on dataset characteristics.  

---

## 2. Practical Implications
- Popularity-based recommendation is highly effective in datasets with strong item popularity bias.  
- Matrix Factorization is more suitable when user-item interactions are denser and more balanced.  
- KNN-based collaborative filtering is sensitive to sparsity and may not perform well without sufficient data.  
- In real-world systems, baseline models like POPULAR remain strong benchmarks and should not be ignored.  

---

## 3. Recommendations
- For production systems with sparse data, start with a **POPULAR baseline** as a strong benchmark.  
- Improve MF performance by tuning hyperparameters (latent factors, regularization, iterations).  
- Consider hybrid recommendation approaches (e.g., combining POPULAR + MF) to improve robustness.  
- Collect more interaction data to improve collaborative filtering performance, especially for KNN-based models.  